\FAQCategory{Documentation}
\FAQquestion{Are there Research reports available?}
\FAQanswer{Yes. Publications and research reports on BALL are listed on our website: 
\FAQURL{http://www.ball-project.org}
}
\FAQquestion{Where do I get the latest version?}
\FAQanswer{The latest version of BALL, bug fixes, and updates are available from our 
website
\FAQURL{http://www.ball-project.org}
}
\FAQquestion{Is there further documentation besides this document?}
\FAQanswer{Further documentation, a Wiki and a bug tracker, can be found on 
\FAQURL{https://github.com/BALL-Project/ball/}
providing a Code Library with Code snippets, further tutorials, Release Notes, and FAQs.
}
\FAQquestion{How do I use BALLView?}
\FAQanswer{Take a look at the built-in BALLView Tutorial or on the documentation at http://www.ball-project.org/Documentation or under BALL/doc/BALLView}
\FAQCategory{Installation}
\FAQquestion{Will BALL run on my hardware/with my compiler?}
\FAQanswer{BALL should run on all major current platforms, in particular Linux, MacOS X, and Windows. Older BALL
versions have been regularly tested also on more exotic platforms, and these should also work; however, the BALL
developers no longer have access to these platforms and cannot test new versions on them.
}
\FAQCategory{License}
\FAQquestion{Under what kind of license is BALL available?}
\FAQanswer{
	BALL is mostly being distributed under the Lesser GNU Public License (LGPL). Parts of BALL (BALLView) are under the GNU Public License (GPL).
}
