\section{Testing the Library}
\index{testing}
\index{test programs}

BALL provides an extensive suite of test programs to ensure the correctness of
the code on all platforms. This test suite requires a lot of patience since
the compilation takes quite some time. However, we recommend to run these
tests to ensure that the library is fully operational. At the moment, the test
suite does not yet cover all functionality of BALL, but only some chosen
classes.  The test programs are located in the directory {\tt
BALL/source/TEST}.  To compile and run the test suite, use "{\tt make test}".
Please make sure that {\tt LD\_LIBRARY\_PATH} is correctly set, otherwise the
execution of the test programs will fail.

Each of the test programs tests one or more classes of BALL. When a test
program (\eg~{\tt Atom\_test}) is run, the program prints either "OK" (if all
tests passed) or "FAILED" if any of the tests failed. "{\tt make test}" runs
all tests and complains if a certain test fails.  If this happens, please let
us know by mailing the files specified in the output of "{\tt make test}" to
one of the developers - this helps us to improve the library.  The mail
address for bug reports is:
\begin{enumerate}
	\item[] \eMail{ball-bugs@mpi-sb.mpg.de}
\end{enumerate}

\noindent
For all bug reports, please include your system configuration, the file
\file{config.log} (which contains the results of tests configure performed),
and the file \file{BALL/include/BALL/config.h} (which contains the compiler
defines used by BALL).

\subsection{Benchmarks}

If you want to know how fast the version of BALL is compared to other systems,
you might want to run the benchmark suite in {\tt BALL/source/BENCHMARKS}.
You can compile the benchmark suite with {\tt make} and then run the
benchmarks with {\tt make bench}. Depending on your hardware and 
whether you compiled BALL with or without optimization, running the benchmarks
will take up to several minutes. Upon termination, it will print an overall
number, the BALLStones. This number is a crude estimate of the performance you
can expect for a mix of typical BALL applications. The benchmark suite
currently includes benchmarks for the BALL kernel data structures, file I/O,
the AMBER force field, and the Poisson-Boltzmann solver. The BALL web site
contains a list of benchmark results for different platforms, please feel free
to submit your results for inclusion.
